---
title: "Portland Housing Market Analysis"
author: "Cormac Dacker, Tani Shalayuy
output: html_notebook
---

# Portland Housing Market Analysis
## Cormac Dacker, Tani Shalayuy

### Introduction
#### Getting housing at an affordable price is harder now more than ever. With the rise of the internet, it is easier to find a house, but it is also easier for people to find the same house. This has led to a persived scramble to find
We wanted to see if there was a way to predict housing prices in Portland, Oregon. We used a dataset from Kaggle that had information on houses sold in Portland from 2021 to 2022. We used this dataset to create a model that could predict housing prices in Portland. We also wanted to see if there was a way to predict the price of a house based on the number of bedrooms, bathrooms, and square footage. We used a dataset from Kaggle that had information on houses sold in Portland from 2014 to 2015. We used this dataset to create a model that could predict housing prices in Portland. We also wanted to see if there was a way to predict the price of a house based on the number of bedrooms, bathrooms, and square footage.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load libraries
library(ggplot2)
library(tidyverse)
library(xkcd)
library(gridExtra)
library(dplyr)


# Load data
root_wd = getwd()
portland_df = read_csv(paste0(root_wd, "portland_housing.csv"))

# Data cleaning and manipulation
portland_df = portland_df %>% rename("Pre and Elementary School Distance" = "schools/0/distance",
                                     "Middle School Distance" = "schools/1/distance",
                                     "High School Distance" = "schools/2/distance",
                                     pre_n_elementary_school_level = "schools/0/level",
                                     middle_school_level = "schools/1/level",
                                     high_school_level = "schools/2/level",
                                     city = "address/city", zipcode = "address/zipcode",
                                     reso_homeType = "resoFacts/homeType")
# convert relevant cat to numeric
portland_df$daysOnZillow = as.numeric(portland_df$daysOnZillow)
portland_df$favoriteCount = as.numeric(portland_df$favoriteCount)

# portland_df_affordable = portland_df %>% filter(price < 1000000)
# portland_df_affordable = portland_df_affordable %>% filter(reso_homeType == "SingleFamily")

school_df = portland_df %>% pivot_longer(cols = c("Pre and Elementary School Distance", "Middle School Distance", "High School Distance"), names_to = "school_type", values_to = "distance_to_school")
# turn daysOnZillow into a factor
school_df$daysOnZillow = as.numeric(school_df$daysOnZillow)

# filter for only sold homes
sold_homes = portland_df %>% filter(homeStatus == "RECENTLY_SOLD")
# create a col that is decade built for sold homes
sold_homes = sold_homes %>% mutate(decadeBuilt = floor(yearBuilt/20)*20)
# convert decadeBuilt to a factor
sold_homes = sold_homes %>% mutate(decadeBuilt = as.factor(decadeBuilt))
# drop houses from before 1900 and NA values
sold_homes = sold_homes %>% filter(yearBuilt > 1920) %>% drop_na(decadeBuilt)

# a df with dateSold converted to a human readable date
date_df = sold_homes %>% mutate(dateSold = dateSold %/% 1000) %>% mutate(dateSold = as.POSIXct(dateSold, origin = "1970-01-01")) %>% mutate(dateSold = as.Date(dateSold))
date_df <- date_df %>%
  mutate(homeType = recode(homeType, "SINGLE_FAMILY" = "Single Family House",
                           "APARTMENT" = "Apartment",  "TOWNHOUSE" = "Townhouse"))
# only keep single family homes, apartments, and townhouses
date_df = date_df %>% filter(homeType == "Single Family House" | homeType == "Apartment" | homeType == "Townhouse")
# create a new col weeksold that is the week of the year the house was sold
date_df = date_df %>% mutate(weeksold = week(dateSold))
# drop the last week of the year becasue brokers will defaul to the last week of the year if they don't know the exact date
date_df = date_df %>% filter(weeksold < 51)
# convert weeksold to factor
date_df = date_df %>% mutate(weeksold = as.factor(weeksold))
# create a new col month sold
date_df = date_df %>% mutate(monthsold = month(dateSold))
# convert monthsold to factor
date_df = date_df %>% mutate(monthsold = as.factor(monthsold))

broker_top = portland_df %>% group_by(brokerageName) %>% drop_na(brokerageName) %>%
  summarize(n = n()) %>%
  arrange(desc(n)) %>% head(10)
```

#### If you're around my age the idea of 'starting a family' becomes an increasingly pressing issue. So when buying your home you might want to think ahead and position yourself close to schools. But we all know that the closer you are to a school the more expensive the housing is... or is that really the case?
```{r}
ggplot(school_df, aes(x = distance_to_school, y = price)) + geom_point(alpha=.2, position = position_jitter(width = 0.04, height = 0)) +
  facet_wrap(~school_type,ncol = 1) +
  labs(title = "Distance to School vs Price", x = "Distance to School", y = "Price") +
  # add a linear regression line
  geom_smooth(method = "lm", se = FALSE) +
  # remove the gab between the x axis and the first tick
  # scale_x_continuous(expand = c(0, 0)) +
  xkcdaxis(yrange = c(0, 1000000), xrange = c(0, 5)) +
  # add font
  theme(text = element_text(family = "xkcd"))
```
#### The most interesting discovery we found blah blah blah

```{r}

xrange = range(min(portland_df$daysOnZillow, na.rm = TRUE), max(portland_df$daysOnZillow, na.rm = TRUE))
yrange = range(min(portland_df$favoriteCount, na.rm = TRUE), max(portland_df$favoriteCount, na.rm = TRUE))
ratioxy = diff(xrange)/diff(yrange)

mapping = aes(x, y, scale, ratioxy, angleofspine, anglerighthumerus, anglelefthumerus,
               anglerightradius, angleleftradius, anglerightleg, angleleftleg, angleofneck,
               linetype=city)
dataman = data.frame(x=150, y=100, scale=30, ratioxy=ratioxy, angleofspine=-pi/2, anglerighthumerus=pi/4, anglelefthumerus=pi/2+pi/4,
                     anglerightradius=2*pi/3, angleleftradius=2*pi/3, anglerightleg=3*pi/2-pi/12, angleleftleg=3*pi/2+pi/12,
                     angleofneck=runif(1, 3*pi/2-pi/10, 3*pi/2+pi/10), city="Liliput")

# 2. Likes/time on zillow and what facets may affect this

xkcd_like = ggplot() + geom_point(aes(daysOnZillow, favoriteCount), data=portland_df, alpha= 0.8) +
  xkcdaxis(xrange,yrange) + xkcdman(mapping, dataman) + labs(title = "Likes vs Days on Zillow", x = "Days on Zillow", y = "Likes") +
# drop the legend
  theme(legend.position = "none") +
# add a line fron the head (150,100) to the text line (170,150)
  annotate("segment", x = 150, xend = 170, y = 120, yend = 150, size = 0.5) +
# add the text "I'm a like bender!" next to the speach line
annotate("text", x = 170, y = 150, label = "I'm a like bender!", size = 10, family = "xkcd", hjust = 0, vjust = 0)
xkcd_like
```

```{r}
xrange = range(min(portland_df$daysOnZillow, na.rm = TRUE), 100)
yrange = range(min(portland_df$favoriteCount, na.rm = TRUE), 100)
# recreate the graph above but lim 100 x and y
xkcd_like2 = portland_df %>% filter(daysOnZillow<=100)%>%ggplot() +
        geom_point(aes(daysOnZillow, favoriteCount), alpha= 0.2, position = position_jitter(width = 0.25, height = 0.25)) +
        xkcdaxis(xrange,yrange) +
        # xkcdman(mapping, dataman) +
        labs(title = "A Closer Look", x = "Days on Zillow", y = "Likes") +
        # drop the legend
        theme(legend.position = "none")
        # add a line fron the head (150,100) to the text line (170,150)
        # annotate("segment", x = 150, xend = 170, y = 120, yend = 150, size = 0.5) +
        # add the text "I'm a like bender!" next to the speach line
        # annotate("text", x = 170, y = 150, label = "I'm a like bender!", size = 10, family = "xkcd", hjust = 0, vjust = 0) + xlim(0, 100) + ylim(0, 100)
xkcd_like2
```
```{r}
grid.arrange(xkcd_like, xkcd_like2, ncol = 2)
```


```{r}
mapping = aes(x, y, scale, ratioxy, angleofspine, anglerighthumerus, anglelefthumerus,
              anglerightradius, angleleftradius, anglerightleg, angleleftleg, angleofneck)
dataman = data.frame(x=c(150,290),
                     y=c(750,900),
                     scale=c(50,50),
                     ratioxy=.5, angleofspine=c(3*pi/2+pi/15, -pi/2),
                     anglerighthumerus=c(pi/4, -pi/6),
                     anglelefthumerus=c(5*pi/6, pi+pi/6),
                     anglerightradius=c(pi/4, -pi/3),
                     angleleftradius=c(2*pi/3, -pi/4),
                     anglerightleg=c(4*pi/3,3*pi/2-pi/12),
                     angleleftleg=c(5*pi/3,-pi/2),
                     angleofneck=runif(1, 3*pi/2-pi/10, 3*pi/2+pi/10))

# plot days on zillow for sold homes, as a histogram
doz_histogram = ggplot(sold_homes, aes(x = daysOnZillow)) +
  geom_histogram(binwidth = 10, color = "black", fill = "white") +
  labs(title = "Days on Zillow for Sold Homes", x = "Days on Zillow", y = "Count") +   xkcdaxis(yrange = c(0, 950), xrange = c(0, 400)) + xkcdman(mapping, dataman) +
  # add font
  theme(text = element_text(family = "xkcd"))+
    # add a geom rect between 100 and 200
  annotate('rect', xmin = 100, xmax=221, ymin=0, ymax=950, alpha=.5, fill = 'grey') +
# add the anotation "help!!"
  annotate("text", x = 170, y = 800, label = "Help!!", size = 5, family = "xkcd", hjust = 0, vjust = 0)

doz_histogram
```
```{r}
grid.arrange(xkcd_like, doz_histogram, ncol = 2)
```



```{r}

# xkcd style
ggplot(sold_homes, aes(x = daysOnZillow)) +
  geom_histogram(binwidth = 10, color = "black", fill = "white") +
  labs(title = "How does age affect prices?", x = "Days on Zillow", y = "Count") +
  xkcdaxis(yrange = c(0,300), xrange = c(0, 400)) +
  # add font
  theme(text = element_text(family = "xkcd")) + facet_wrap(~decadeBuilt)
```
```{r}
library(plotly)
# make a scatter plot of price vs avg living area with days on zillow as the frame
# not in xkdc style
# ggplotly(ggplot(sold_homes, aes(x = livingArea, y = price, frame = daysOnZillow)) +
#   geom_point(alpha = 0.5) +
#   labs(title = "Price vs Living Area", x = "Living Area", y = "Price"))

# do one dot what displays avg living area and avg price for each day on zillow
ggplotly(date_df %>% group_by(weeksold) %>% summarize(avg_livingArea = mean(livingArea), avg_price = mean(price), avg_lotSize = mean(lotSize), avg_bathrooms = mean(bathrooms), avg_bedrooms = mean(bedrooms), avg_daysOnZillow = mean(daysOnZillow), avg_favoriteCount = mean(favoriteCount), avg_yearBuilt = mean(yearBuilt)) %>%
  ggplot(aes(x = avg_livingArea, y = avg_price, frame = weeksold, size = avg_lotSize, color=avg_yearBuilt)) +
  geom_point(alpha = 0.5) +
  labs(title = "Price vs Living Area", x = "Living Area", y = "Price"))


```




```{r}
xrange = range(0, 10)
yrange = range(0, 1600)
ratioxy = diff(xrange)/diff(yrange)

mapping = aes(x, y, scale, ratioxy, angleofspine, anglerighthumerus, anglelefthumerus,
              anglerightradius, angleleftradius, anglerightleg, angleleftleg, angleofneck)
dataman = data.frame(x=c(.38,.93),
                         y=c(500,800),
                     scale=c(100,100),
                     ratioxy=ratioxy, angleofspine=c(3*pi/2+pi/15, -pi/2),
                     anglerighthumerus=c(pi/4, -pi/6),
                     anglelefthumerus=c(pi/6, pi+pi/6),
                     anglerightradius=c(pi/3, -pi/3),
                     angleleftradius=c(pi/4, -pi/4),
                     anglerightleg=3*pi/2-pi/12,
                     angleleftleg=-pi/2,
                     angleofneck=runif(1, 3*pi/2-pi/10, 3*pi/2+pi/10))
# xkcd style

ggplot(broker_top, aes(x = reorder(brokerageName, n), y = n)) +
  geom_bar(stat = "identity") +
  labs(title = "Where Should You Go to Find a House in Portland?", x = "Brokerage", y = "Number of Houses Listed") +
  xkcdaxis(yrange = c(0, 1600), xrange = c(0, 10)) + xkcdman(mapping, dataman) +
  # rotate the x axis labels
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  # add the line
  # add font
  theme(text = element_text(family = "xkcd")) + theme(plot.title = element_text(size = 28)) +
 # add a text anotation 'Scar?! Brother?! Help me!'
  annotate("text", x = 1.5, y = 1000, label = "Scar! Brother, Help me!", size = 7, family = "xkcd") +
# add another text anotation 'Long live the king!'
  annotate("text", x = 2.5, y = 800, label = "Long live the king!", size = 7, family = "xkcd")+
  # xkcdline(aes(xbegin=xbegin,ybegin=ybegin,xend=xend,yend=yend), datalines, xjitteramount = 0.12)
# add a line from the text to the bar
  annotate("segment", x = .5, xend = .4, y = 950, yend = 600, size = 0.5)
```












```{r}
house <- portland_df %>%
  mutate(homeType = recode(homeType, "SINGLE_FAMILY" = "Single Family House",
                           "APARTMENT" = "Apartment",  "TOWNHOUSE" = "Townhouse"))
# only keep single family homes, apartments, and townhouses
house <- house %>% filter(homeType == "Single Family House" | homeType == "Apartment" | homeType == "Townhouse")
mapping <- aes(x, y, scale, ratioxy, angleofspine, anglerighthumerus, anglelefthumerus,
               anglerightradius, angleleftradius, anglerightleg, angleleftleg, angleofneck)
dataman <- data.frame(x= c(3000, 5000), y=c(1.5, 1.5), scale = c(30,51) ,ratioxy = ratioxy,
                      angleofspine = -pi/2 , anglerighthumerus = c(pi/4, -pi/6), anglelefthumerus = c(pi/2 + pi/4, pi +pi/6),
                      anglerightradius = c(pi/3, -pi/3), angleleftradius = c(pi/3, -pi/3), anglerightleg = 3*pi/2 - pi / 12,
                      angleleftleg = 3*pi/2 + pi / 12 , angleofneck = runif(1, 3*pi/2-pi/10, 3*pi/2+pi/10))

# xkcd style
lot_vs_price = ggplot(house, aes(x = lotSize, y = price / 1e6)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "What are your thoughts on Lot Size and Price?",
    x = "Lot Size in Square Feet",
    y = "Price in Millions"
  ) +
  xlim(0, 6000) +
  ylim(0, 2) +
  theme_bw() +
  geom_smooth(method = lm, se = FALSE) +
  facet_wrap(~homeType, scales = "free_y") +  # or facet_grid(~homeType)
  theme(strip.text = element_text(face = "bold")) + xkcdaxis(yrange = c(0, 2), xrange = c(0, 6000)) + xkcdman(mapping, dataman) +
  # add font
  theme(text = element_text(family = "xkcd"))
```
```{r}
living_vs_price = ggplot(house, aes(x = livingArea, y = price / 1e6)) +
        geom_point(alpha = 0.5) +
        labs(
                title = "Do you think paying More for a house ensures more Living Space?",
                x = "Living Area in Square Feet",
                y = "Price in Millions"
        ) +
        xlim(0, 6000) +
        ylim(0, 2) +
        theme_bw() +
        geom_smooth(method = lm, se = FALSE) +
        facet_wrap(~homeType, scales = "free_y") +  # or facet_grid(~homeType)
        theme(strip.text = element_text(face = "bold")) + xkcdaxis(yrange = c(0, 2), xrange = c(0, 6000)) +
        # xkcdman(mapping, dataman) +
        # add font
        theme(text = element_text(family = "xkcd"))
```
```{r}
grid.arrange(lot_vs_price, living_vs_price, ncol = 1)
```
```{r}

# lets use a violin plot instead
# ggplot(date_df, aes(x = monthsold, y = price)) +
#   geom_violin() +
#   labs(title = "Average Price of Sold Homes by Week", x = "Month", y = "Price") +
#   # add font
#   theme(text = element_text(family = "xkcd")) + facet_wrap(~homeType) + theme_bw()

# convert to fully xkcd style
ggplot(date_df, aes(x = monthsold, y = price)) +
  geom_violin(fill = "white", color = "black") +
  labs(title = "Average Price of Sold Homes by Month", x = "Month", y = "Price") +
  # add font
  facet_wrap(~homeType) + theme(strip.text = element_text(face = "bold"))+
  theme_bw() +
  theme(text = element_text(family = "xkcd"))
  xkcdaxis(yrange = c(0, 1800000), xrange = c(0, 13))
```
```{r}

```
```{r}

```

```{r}

```
